---
title: "Fisheries  Stability"
author: "Julia L. Blanchard"
date: "25/08/2021"
output: html_document
---

Fisheries stability under multiple drivers of change. The main goal of this model is to assess whether/how changes in fisheries development/maagement (e.g effort, TACs) can prevent collapse (or help recover) potential exploitable fish biomass.

We use a  trait-based modelling approach combined with climate scenario and reconstructed  catch and effort data, for the Southeast Australian Shelf Large Marine Ecosystem (LME 42). At a later stage, this work will be extended to all LMEs. This work is a contribtion to FishMIP and co-funded by Dalhousie University.

First, let's have a look at the catch and effort, Yannick's previous data. Hopefully this will be updated soon once we have more detail  on the  revised mapping methods - which need to not rely on catch data (need to check with YR whether the catch and effort data used here are actually independent, and if so how they are linked).
```{r}
library(tidyverse)
remotes::install_github("sizespectrum/mizerExperimental")
library(mizerExperimental)
#library(mizerMR)
library(tibble)

```

```{r,eval=FALSE}

catch<-read.csv("../data/Catch_PelDem_by_FGroup_AreaV4.csv")
#check names
names(catch)

# subset LME 42 and FCounty == AUS and Sector == Industrial & FGroup == Large pelagics
catch<-catch[catch$LMEnbr==42,]
#catch<-catch[catch$Fcountry=="AUS",]
#catch<-catch[catch$Sector=="Industrial",]
#catch<-catch[catch$FGroup=="pelagic>=90cm",]
catch
# When I subset for one FGroup & AUS only I expected this subset would produce one single time series - what are the extra points for? # WHY ARE THERE NOT  AN EQUAL NUMBER OF CELLS IN EACH ROW? HOW WAS THIS AGGREGATED? ALSO WHY DIFFERENT FAOnbrs (not FAO areas??) What does X stand for ??

t <- ggplot(catch, aes(x=Year, y=Reported,color=as.factor(X))) + geom_point()
t

### the data are not clear enough to me - I need the total catch only in this LME per FG - are this grid cells?? How were the cells allocated to LMEs - there seems to be a mapping issue? YR could you put the code that allocates your country data to spatial grid cells and then re-aggregates to these LMEs on github please so we can understand and repeat the methods.

# I assume it  is OK to sum all rows by LME here.
  
catch_sum<-aggregate(catch$Reported,list(catch$Year,catch$FGroup),sum)
names(catch_sum)<-c("Year","FGroup","Reported")

catch_sum_IUUs<-aggregate(catch$IUUs,list(catch$Year,catch$FGroup),sum)
names(catch_sum_IUUs)<-c("Year","FGroup","IUUs")

catch_sum$catch<-catch_sum_IUUs$IUUs + catch_sum$Reported

# stacked area chart

# Library
library(viridis)
library(hrbrthemes)
# Plot
ggplot(catch_sum, aes(x=Year, y=catch, fill=as.factor(FGroup))) + 
    geom_area() +
    scale_fill_viridis(discrete = T) +
    theme_ipsum()



effort<-read.csv("../data/Effort_PelDem_by_FGroup_AreaV4.csv")
#check names
names(effort)
# subset LME 42 and FCounty == AUS and Sector == Industrial & FGroup == Large pelagics
effort<-effort[effort$LMEnbr==42,]
effort<-effort[effort$Fcountry=="AUS",]
#effort<-effort[effort$Sector=="Industrial",]
#effort<-effort[effort$FGroup=="pelagic>=90cm",]
effort


effort_sum<-aggregate(effort$NomEffReported,list(effort$Year,effort$Sector,effort$FGroup),sum)
names(effort_sum)<-c("Year","Sector", "FGroup","NomEffReported")

effort_sum_IUUs<-aggregate(effort$NomEffIUU,list(effort$Year,effort$Sector,effort$FGroup),sum)
names(effort_sum_IUUs)<-c("Year","Sector","FGroup","NomEffIUU")

effort_sum$effort<-effort_sum_IUUs$NomEffIUU + effort_sum$NomEffReported

# Plot
ggplot(effort_sum, aes(x=Year, y=effort, fill=FGroup)) + 
    geom_area() +
     scale_fill_viridis(discrete = T) 

 #   scale_y_continuous(trans='log10')
 #   theme_ipsum()

# the data look a little weird to me... why so many values close to  zero?
```


Let's have a look at SAUP catches instead.....

```{r}
# Read in SAUP catches instead
SAUPcatches<-read.csv("../data/SAU LME 42 v48-0/SAU LME 42 v48-0.csv")

head(SAUPcatches)


# also from SAUP for this LME:
#LME area	1,177,870 km2
#Shelf Area	219,772 km2
#Inshore Fishing Area (IFA)	194,195 km2
#Tropical Coral Reefs	0 % of world
#Seamounts	0.168 % of world
#Primary production	514.76 mgCm-2day-1

catch_sum2<-aggregate(SAUPcatches$tonnes,list(SAUPcatches$year,SAUPcatches$functional_group),sum)
names(catch_sum2)<-c("year","functional_group","tonnes")

# Plot
ggplot(catch_sum2, aes(year, y=tonnes, fill=functional_group )) + geom_area() 

# check relative dominances of FGs in one year
c2010<-catch_sum2[catch_sum2$year==2010,]
c2010[order(c2010$tonnes,decreasing = T),]

# extract only fishMIP FGs

unique(catch_sum2$functional_group)
#unique(effort_sum$FGroup)

#unique(effort_sum$FGroup)
#[1] "demersal<30cm"   "demersal>=90cm"  "demersal30-90cm"
#[4] "pelagic<30cm"    "pelagic>=90cm"   "pelagic30-90cm" 

fishmip_groups<-c("Small demersals (<30 cm)","Medium demersals (30 - 89 cm)", 
"Large demersals (>=90 cm)", 
"Small pelagics (<30 cm)",
"Medium pelagics (30 - 89 cm)",
"Large pelagics (>=90 cm)")

catch<-filter(catch_sum2, functional_group %in% fishmip_groups)
names(catch)[3]<-"value"
# Plot
ggplot(catch, aes(year, y=value, fill=functional_group )) + geom_area() 

```


```{r}

library(purrr)
library(zoo)
library(Hmisc)
library(mgcv)
library(data.table)
library(tidyverse)
library(ggpubr)
library(rvest)
library(stringr)
library(magrittr)
library(Kendall)
library(boot)

```

Function to run the EWS metrics


```{r}
runEWS<-function(results=results) {
 window <- c(0.3, 0.4,0.5, 0.6,0.7)

for(i in window){
  
  assign(paste0("window", i), 
    results %>%
   # filter(variable=="catch") %>%
   # filter(sim == 1) %>% 
    group_by(year) %>% 
    summarise(value=sum(value)) %>%
    mutate(Normalized_value = (value - mean(value))/sd(value)) %>% 
    mutate(log_Value = log10(value)) %>% 
    #mutate(Element="Fisheries") %>% 
    mutate(Differenced = (resid(gam(value~s(year))))) %>% #diag(1:4)
  #  mutate(Differenced = c(0, diff(value))) %>% ### fit a gam value smoothed by year and then get residuals and then run everything below, call it Difference
    mutate(Variance = rollapply(data = Differenced, 
                                width = i*length(Differenced), 
                                FUN = sd, 
                                align = "right", 
                                fill = NA, 
                                na.rm = T)) %>% 
    mutate(Mean = rollapply(data = Differenced, 
                            width = i*length(Differenced), 
                            FUN = mean, 
                            align = "right", 
                            fill = NA, 
                            na.rm = T)) %>%
   # mutate(CV=Variance/Mean) %>%
    mutate(AR1 = rollapply(data = (Differenced), 
                           width = i*length(Differenced), 
                           FUN = function(z,  na.rm = FALSE){
                             return(acf(z, plot=FALSE)$acf[2])
                           },  
                           align = "right", 
                           fill = NA, 
                           na.rm = T)) %>% 
    mutate(Return_rate = rollapply(data=Differenced,
                                   width= i*length(Differenced),
                                   FUN = function(z, na.rm=FALSE){
                                     ar1 <- ar.ols(z,demean=FALSE, order.max = 1, aic = FALSE, intercept = FALSE)
                                     return(return_rate = 1/ar1$ar[1])
                                   },
                                   align = "right", 
                                   fill = NA, 
                                   na.rm = T)) %>% 
    mutate(Kurtosis = rollapply(data=Differenced,
                                   width=i*length(Differenced),
                                   FUN = moments::kurtosis,
                                   align = "right", 
                                   fill = NA, 
                                   na.rm = T)) %>% 
    mutate(Skewness = rollapply(data=Differenced,
                                   width=i*length(Differenced),
                                   FUN = moments::skewness,
                                   align = "right", 
                                   fill = NA, 
                                   na.rm = T)) %>% 
    mutate(Width = i)%>% 

    # still good. used to create a composite metric    
    mutate(Diff_AR1= rollapply(data = Differenced, 
                           width =i* length(Differenced), 
                           FUN = function(z,  na.rm = FALSE){
                             return(acf(z, plot=FALSE)$acf[2])
                           },  
                           align = "right", 
                           fill = NA, 
                           na.rm = T)))
  
}

#fisheries <- bind_rows(window0.3, window0.4, window0.5, window0.6,window0.7)
fisheries <- bind_rows(window0.3, window0.4, window0.5, window0.6,window0.7)

fisheries
}

fisheries<-runEWS(catch)


```


Plot the times series of the datae

```{r}
(landings <- ggplot(fisheries %>% filter(Width ==0.5),
       aes(x=year)) +
    geom_line(aes(y=value), colour="#172869", size=1.1)+
    theme_pubr()+
    labs(y=bquote(Tonnes)))


(residuals <- ggplot(fisheries %>% filter(Width ==0.5),
       aes(x=year)) +
  geom_line(aes(y=Differenced), colour="#EA7580", size=1.1)+
    theme_pubr()+
    theme(
      text = element_text(size=9),
      axis.title.x = element_blank()
    )+
    geom_hline(yintercept = 0, linetype=1, colour="grey90", size=0.75)+
    scale_x_continuous(breaks=c(1960,1980,2000,2020), labels = c("1960","1980","2000","2020"), limits = c(1950,2020))+
    labs(y=bquote(Tonnes~(10^3)))
)


(landings_residuals <- ggarrange(landings, residuals, 
          labels = c("a", "b"),
          font.label = list(size=9),
          ncol = 2,
          nrow = 1))

```


Calculate EWS

```{r}

variance_ts <- fisheries$Variance[!is.na(fisheries$Variance)]

library(Kendall)
# trend in Variance - positive then increase in variance
 (variance_tau <- Kendall(1:length(variance_ts), variance_ts)$tau[1])

 tau.fc<-function(x,y){
    y<-sample(x, replace = T)
   tau<-as.numeric(Kendall(1:length(y), y)$tau)
   return(tau)
 }

 #don't worry about  - pavlue for plots
 boot_mod <- boot(data=variance_ts, statistic=tau.fc,  R=1000)
 hist(boot_mod$t, xlim = c(-1, 1))
 abline(v=variance_tau, lty=2, col="red")
# length(which(boot$t>variance_tau))/1000

#reliability investigation into trend test

  vars <- ggplot(data=fisheries, aes(x=year, y=Variance))+
    geom_line()+
    facet_grid(.~Width)+
    theme_pubr()+
      theme(
        text = element_text(size=9),
        axis.title.x = element_blank()
      )+
        scale_y_continuous(breaks = c(8e+3 ,10e+3, 12e+3, 14e+3,16e+3), labels=c("8", "10", "12", "14", "16"))+
      scale_x_continuous(breaks=c(1960,1980,2000,2020), labels = c("1960","1980","2000","2020"), limits = c(1950,2020))+
      labs(y=bquote(Variance~(Tx10^3)))
    
  
  ar1 <- ggplot(data=fisheries, aes(x=year, y=AR1))+
    geom_line()+
    facet_grid(.~Width)+
    theme_pubr()+
      theme(
        text = element_text(size=9),
        axis.title.x = element_blank()
      )+
         scale_x_continuous(breaks=c(1960,1980,2000,2020), labels = c("1960","1980","2000","2020"), limits = c(1950,2020))+
      labs(y="AR1", x="Year")
  
  # add compposite metrcis form Chris's paper:
  
  comp_fisheries <- fisheries %>% 
    drop_na() %>% 
    group_by(Width) %>% 
    nest() %>% 
    mutate(Normalised_variance = map(data, ~((.$Variance-mean(.$Variance))/sd(.$Variance))),
           Normalised_AR = map(data, ~((.$AR1-mean(.$AR1))/sd(.$AR1)))) %>% 
    unnest() %>% 
    mutate(Composite_index = Normalised_variance+Normalised_AR)
  
  
   comp_index <- ggplot(data=comp_fisheries, aes(x=year, y=Composite_index))+
    geom_line()+
    facet_grid(.~Width)+
    theme_pubr()+
      theme(
        text = element_text(size=9),
        axis.title.x = element_blank()
      )+
         scale_x_continuous(breaks=c(1960,1980,2000,2020), labels = c("1960","1980","2000","2020"), limits = c(1950,2020))+
      labs(y=bquote(Composite~index))
   
   
   ggarrange(landings_residuals, vars,ar1, comp_index, 
           nrow=4, ncol=1,
           labels = c("", "c", "d", "e"),
           font.label = list(size=9),
           heights = c(1,0.75,0.75,0.75))
    

st_dev <- sd((fisheries %>% filter(Width==0.5))$Differenced)
 

#bds.test((fisheries %>% filter(Width==0.5))$Differenced, eps = c(0.5*st_dev, 0.75*st_dev, 1*st_dev))


fisheries_test <- fisheries %>% filter(Width==0.5)
qqnorm((fisheries_test$Differenced[2:68]))
qqline(fisheries_test$Differenced[2:68])

```


Now set up a mizer model with these functional groups. Will add multiple resources (bethic and pelagic) later.

```{r}
# check if start  from scratch
speciesParams<-read.csv("../data/fg_params.csv")
inter<-read.csv("../data/inter.csv")[,-1]
params <- newMultispeciesParams(speciesParams, kappa = 1e20,inter)

species_params(params)$R_max<-params@resource_params$kappa*params@species_params$w_inf^-1.5
params@species_params$erepro<-0.01
params@species_params$h<-80
sim<- project(params, t_max = 200, effort = 0)
plot(sim)

#tuneParams(params) - made adjustments to kappa in tueParams and saved file below:

params<-readRDS("../data/params.rds")
sim<- project(params, t_max = 200, effort = 0)
plot(sim)
#tuneGrowth(params)


# by default mizer sets up a gear_params object
params@gear_params
# when effort is set  to > 0 the fishing gets switched on

# have a look:
sim_f<- project(params, t_max = 200, effort = 0.2)
plot(sim_f)
```

Now set up effort through time

```{r}
# set up fishing moratlity (here called effort) time series
years<-1850:2050
species<-speciesParams$species
gear<-"knife_edge_gear"
#effort_input<-matrix(0,nrow=201,ncol=6,dimnames = list(years,species))
effort_input<-matrix(0,nrow=201,ncol=1,dimnames = list(years,gear))
row.names(matrix)

# Historical effort (here we use fishing mortality rate, could separate catchabilty and effort later)

# Use double logistic curve (fishing mortality through time), with a descending limb
# starting from 1850

# f(y) ~ F/(1+ exp(-y0(y-y1))) * F/(1 + exp(-y2(y-y3))
# where:
# y1, the year value of the sigmoid's midpoint;
# F, the curve's maximum value;
# k, the logistic growth rate or steepness of the curve

c1=0.05
c2=0.05 # 0.1 fast, 0.03 medium-MSY, 0.01 slow
yr50a=1980
yr50b=2005
Fmax1=1
Fmax2=1
effort_input[,"knife_edge_gear"]<-(Fmax1/(1+ exp(-c1*(years-yr50a))))*(Fmax2/(1 + exp(c2*(years-yr50b))))
plot(years,effort_input[,"knife_edge_gear"],typ="l")


simt <- project(params, effort = effort_input,initial_n = sim@n[200,,], initial_npp = sim@n[200,])

# Each species yields through time
plotYieldGear(simt)




```

We need to set  up a stochastic model to examine the detectibility of EWS indicators. 
Here we ass environmental red noise (from Gustav's RPubs example). Later we will set up a climate run that takes the mean trend, seasonal and interannual variability (similar to in Romain's final thesis chapter) from plankton and temperature variables.


### Adding  noise to the resource through time

We add noise to the resource function. To prepare for random changes in plankton carrying capacity every half year,we create an environment to maintain state between function calls.
```{r,code_folding = show}
plankton_state <- new.env(parent = emptyenv())
plankton_state$time <- 0
plankton_state$factor <- 1
plankton_state$phi <- 0
plankton_state$sigma <- 0.5

resource_semichemostat_noise<-function (params, n, n_pp, n_other, rates, t, dt, resource_rate, resource_capacity, ...) 
{  plankton_state$time <- plankton_state$time + dt
   # red noise
   #plankton_state$factor <- plankton_state$factor ^ plankton_state$phi* 
          #exp(rnorm(1, 0, plankton_state$sigma))
   # plankton_state$factor <- exp(runif(1, log(1/2), log(2)))
    mur <- resource_rate + rates$resource_mort
   # n_steady <- resource_rate * resource_capacity/plankton_state$factor/mur
     n_steady <- resource_rate * resource_capacity/mur
    n_pp_new <- n_steady + (n_pp - n_steady) * exp(-mur * dt)
    sel <- mur == 0
    n_pp_new[sel] <- n_pp[sel]*exp(runif(1, log(1/2), log(2)))
    n_pp_new
}

```

Now set up model resource_params and carry out a bunc of runs 
```{r}

run_sim<-function(params,num_sims=10,times=1850:2050){
  
 species<-params@species_params$species

 #set up  arrays to store outputs 
 
 biomass_array<-array(NA, dim = c(length(times),length(species) ,num_sims), dimnames = list(times, species, 1:num_sims))
 
 yield_array<-array(NA, dim = c(length(times),length(species) ,num_sims), dimnames = list(times, species, 1:num_sims))
 
params_s<-setResource(params, resource_dynamics = "resource_semichemostat_noise")
   
for(i  in 1:num_sims) {

simt_s <- project(params_s, effort = effort_input,initial_n = sim@n[200,,], initial_npp = i)

 biomass_array[,,i]<-getBiomass(simt_s)
 yield_array[,,i]<-getYield(simt_s)
}

variable<-c(rep("biomass",num_sims*length(times)*length(species)),rep("yield",num_sims*length(times)*length(species)))

result<-rbind(melt(biomass_array),melt(yield_array))

result$variable<-variable

names(result)<-c("year","species","sim","value","variable")

return(result)
}

results<-run_sim(params)

saveRDS(results,"results_sim.rds")

biomass<- results %>% filter(variable=="biomass",sim==1) %>% group_by(year)  %>%summarise(value=sum(value))

yield<- results %>% filter(variable=="yield",sim==1) %>% group_by(year)  %>%summarise(value=sum(value))

# plot(biomass)
# plot(yield)


# Plot
ggplot(results[results$variable=="biomass",], aes(year, y=value, color=species )) + geom_line() 

ggplot(results[results$variable=="yield",], aes(year, y=value, color=species )) + geom_line() 

ggplot(results[results$variable=="yield" & results$sim==1,], aes(year, y=value, fill=species )) + geom_area() 

ggplot(results[results$variable=="yield" & results$sim==1,], aes(year, y=value, fill=species )) + geom_area() 


```








Next  we want to ask the question whether we can predict collapse using EWS - here simply based on the catch...

Need to know year of collapse for each run.

Then when transitions detected.

Need to enable noise to be varied above and vary the rate of fishing development ...


We can see this causes variability  in the times series - which is what we want.


Simply ask is variance rising - and how does biomass and catch look? - and does it decline when management kicks in ? 
3 scenarios?
Add polygons for stoch simulations


```



DO WE SEE A PATTER IN THE DATA??

If you detect in the landings when the the biomass collapse happenss but NOT detected in the biomass EWS.
Biomass indicator - B/B0 - find the parameter that generate the early warning signals???
What needs to be true for 

May not be a feature of the biological systems but of the social-ecological system
How accurately we are hunting affects things....



NOW Repeat above but with SAUP catches

```{r}
fisheries_mod_bio<-runEWS(variable="biomass")
fisheries_mod_yield<-runEWS(variable="yield")

```

Try plotting this

```{r}
(landings <- ggplot(obs_fisheries %>% filter(Width ==0.5), aes(x=year)) +
  geom_line(aes(y=value), colour="#172869", size=1.1)+
 #geom_line(aes(y=predict(gam(Value~s(Year)))), size=0.6, colour="grey60")+
    theme_pubr()+    theme(text = element_text(size=9), axis.title.x = element_blank() +
      #plot.margin = unit(c(t=0.5, r=0.1,b=0.1,l=0.1), "cm"))+
   # geom_vline(xintercept = 2007, linetype=2, colour="grey80", size=0.75)+
#    scale_y_continuous(breaks = c(0 ,0.5e+5, 1e+5, 1.5e+5,2e+5, 2.5e+5), labels=c("0", "50", "100", "150", "200", "250"), limits = c(0, 2.7e+5))+
 #   scale_x_continuous(breaks=c(1960,1980,2000,2020), labels = c("1960","1980","2000","2020"), limits = c(1950,2020))+
    labs(y=bquote(Tonnes))+
    #annotation_custom(fishing_img, xmin = 1950, 1960, ymin = 1.9e+5, ymax=2.9e+5)+
    annotate("text", x=1950, y=30,label="Fisheries landings", hjust=0, size=2.9)
#+
#  ggsave("EWS_landings.tiff", dpi=600, device="tiff", width = 5.5, height=4, units = #"cm")

)


(residuals <- ggplot(obs_fisheries %>% filter(Width ==0.5),
       aes(x=year)) +
  geom_line(aes(y=Differenced), colour="#EA7580", size=1.1)+
  # geom_line(aes(y=Mean), colour="#EA7580", size=1.1)+
  #geom_line(aes(y=predict(gam(Value~s(Year)))), size=0.6, colour="grey60")+
    theme_pubr()+
    theme(
      text = element_text(size=9),
      axis.title.x = element_blank()
      #plot.margin = unit(c(t=0.5, r=0.1,b=0.1,l=0.1), "cm")
      
      
    )+
    geom_hline(yintercept = 0, linetype=1, colour="grey90", size=0.75)+
 #   geom_vline(xintercept = 2007, linetype=2, colour="grey80", size=0.75)+
    #scale_y_continuous(breaks = c(0 ,0.5e+5, 1e+5, 1.5e+5,2e+5, 2.5e+5), labels=c("0", "50", "100", "150", "200", "250"), limits = c(0, 2.7e+5))+
    scale_x_continuous(breaks=c(1960,1980,2000,2020), labels = c("1960","1980","2000","2020"), limits = c(1950,2020))+
    labs(y=bquote(Tonnes))
    #annotation_custom(fishing_img, xmin = 1950, 1960, ymin = 1.9e+5, ymax=2.9e+5)+
  # annotate("text", x=1950, y=39e+3,label="First-difference detrended", hjust=0, size=2.9)
  #+
#    ggsave("EWS_differenced_landings.tiff", dpi=600, device="tiff", width = 5.5, height=4, units = "cm")

)


(landings_residuals <- ggarrange(landings, residuals, 
          labels = c("a", "b"),
          font.label = list(size=9),
          ncol = 2,
          nrow = 1))

```



## To add later: Multiple resources  - how to set up the reosurce_dynamics here? Or can they only be semi-chemostat?
```{r}
# library(mizerMR)
# library(tibble)
# resource_params <- tribble(
#     ~resource,  ~kappa, ~lambda, ~r_pp, ~w_min, ~w_max,
#  "Resource 1",    1e11,    1.05,   10,   1e-12 ,  NA,
#  "Resource 2",    1e9,    1.05,   10,   NA ,   100)
# 
# 
# resource_interaction <- matrix(runif(6*2), nrow = 6, ncol = 2)
# resource_interaction[,1] <-c(1,0.5,1,0.5, 1,0.5)
# resource_interaction[,2] <-c(0,0.5,0,0.5, 0,0.5)
# 
# 
# params_sMR <- setMultipleResources(params_s, resource_params) 
# sim_mr <- project(params_sMR, t_max = 200, t_save = 0.2,effort = 0)
# 
# plotSpectra(sim_mr)

```

